Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-10-05 paper/manuscript.tex
Mainly updates to software package section from @IAlibay
Also minor tweaks from @apbhati

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of hydration or a free energy of binding. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, [MENTION OTHERS TOO], etc., circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicates to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
If using a large number of replicates these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
Another approach, known as integration over parts~\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates~\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}.
A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

In addition to statistical uncertainty and sampling, a variety of other factors can impact results from binding free energy calculations. In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecule(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices. Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
Incidentally, Vassaux \textit{et al.} perfor med a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}. They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.

@@

	\item \href{http://getyank.org/latest/index.html}{YANK} is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
        \item \href{https://github.com/choderalab/openmmtools} provides tooling for creating alchemical simulations using OpenMM. These include tools for alchemically modifying OpenMM Systems, running multistate simulations and adding alchemical restraints.

@@

    \item \href{https://openfree.energy/}{Open Free Energy} is a platform which offers and allows the creations of alchemical simulation workflows for both absolute and relative binding free energies. Current offerings, notably include a Perses-based \url{https://github.com/choderalab/perses/tree/main/perses} hybrid topology method using OpenMM.

@@

    Konnektor: Tooling to aid in the creation of alchemical perrturbation networks~\cite{}.

@@

    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM.

@@

	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}~\cite{daviddotson2020alchemistry}.
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper      
	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
	\item Cinnabar: Standardising alchemical free energy analysis. Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \url{https://github.com/OpenFreeEnergy/cinnabar}.
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages 
        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit} FE-Toolkit: collection of packages for free energy analysis, including network-wide MBAR analysis~/cite{}.
    \url{https://github.com/michellab/freenrgworkflows}.
    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
	\end{itemize}
\item[] Generally, commercial software will offer more complete pipelines in which standalone analysis applications are not necessarily needed; free and open source packages often require manual analysis but allow more flexibility and modification.
\end{itemize}

>>>>

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\section{Simulation prerequisites}
\label{sec:prerequisites}
Alchemical free energy protocols as discussed below (Sec.~\ref{sec:simulation_protocol_choice}) are defined for a specific type of free energy calculation, i.e. a free energy of solvation or a free energy of binding. Different types of simulations require different choices for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimized structure of a solute can be generated with a simple tool such as Open Babel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
Additionally, SCDT approaches, such as TIES~\cite{bhati2017}, circumvent this issue by design as they do not need any restraints to hold MCSS together between the two end-points.

@@

There are multiple strategies to alleviate such sampling challenges in an alchemical calculation. A straightforward but computationally expensive method is to run longer simulations with multiple replicas to increase conformational heterogeneity~\cite{suruzhon2021sensitivity}.
If using a large number of replicas these are often termed ensemble approaches and an example of such a sampling scheme is the TIES method~\cite{bhati2017} developed by Bhati et al., where a minimum of 5 to 10 replicas are recommended to make up the ensemble for RBFE and ABFE calculations respectively~\cite{bhati2022, bhati2025}.
Another approach, known as integration over parts~\cite{jayachandran2006parallelized}, involves performing separate alchemical calculations limited to a series of binding macrostates and  then computing the overall free energy difference using the free energy combination formula~\cite{gallicchio2011recent,hsu2023alchemical}, i.e., computing the negative log-sum-exp of the Boltzmann weights from each macrostate. In the integration over parts approach, restraining potentials can be used to restrict the sampling to specific macrostates~\cite{khuttan2023taming}. However, in practice, restraining potentials are sometimes omitted when the states are known to be separated by large free energy barriers and do not interconvert within the simulation time scale, such as cases like Fig.~\ref{fig:configurational_sampling}\textbf{B}.
A generally more effective strategy is to bias relevant configurational collective variables (CVs) during the alchemical simulation. This can be achieved by enhancing sampling within individual alchemical states and swapping conformations between alchemical states (preferably from states separated by large free energy barriers) via replica exchange protocols. More recently, several approaches have been proposed to simultaneously bias both configurational CV and alchemical variables, which can effectively address both scenarios Fig.~\ref{fig:configurational_sampling}\textbf{B} and \textbf{C}. Examples include alchemical metadynamics~\cite{hsu2023alchemical}, $\lambda$-ABF (adaptive biasing force)~\cite{lagardere2024lambda}, and optimized alchemical enhanced sampling (ACES)~\cite{lee2023aces}  and approaches that integrate well-tempered metadynamics~\cite{barducci2008well} with alchemical methods~\cite{zhou2025zooming, khuttan2024make}, to name a few.

@@

In addition to statistical uncertainty and sampling, a variety of other factors can impact results from binding free energy calculations. In addition to the choice of initial configuration, results can depend on the choice of force field for the protein/receptor, water, and small molecule(s), so rerunning calculations with different choices of force field can also be used to assess how sensitive results and conclusions are to these particular choices. Other factors, like system preparation (choice of protonation state, tautomer, counterion presence, salt concentration, etc.) can also substantially impact results~\cite{mobley2017predicting, mobley2017predictinga}, so unless modelers are confident they have these factors correct, sensitivity to these choices may also need to be examined.
Incidentally, Vassaux \textit{et al.} performed a systematic study comparing the effect of both aleatoric as well as parametric uncertainties for protein-ligand free energy calculations using NAMD~\cite{vassaux2021}. They found that, indeed, ensemble simulations are necessary to control both sources of uncertainties arising in simulations, just as is anticipated in the field of uncertainty quantification.

@@

        \item \href{https://github.com/choderalab/openmmtools}{OpenMMTools} provides tooling for creating alchemical simulations using OpenMM. These include tools for alchemically modifying OpenMM Systems, running multistate simulations and adding alchemical restraints \cite{john_chodera_2025_17238654}.

@@

    \item \href{https://openfree.energy/}{Open Free Energy} is a platform which offers and allows the creations of alchemical simulation workflows for both absolute and relative binding free energies. Current offerings, notably includes a \href{https://github.com/choderalab/perses/tree/main/perses}{Perses-based} hybrid topology method using OpenMM \cite{alibay_2025_17258732}.
    \item \href{https://github.com/Gallicchio-Lab/AToM-OpenMM}{AToM-OpenMM} is a Python package for running simulations using the Alchemical Transfer Method in OpenMM \cite{azimi2022relative, wu2021alchemical,gallicchio2008,gallicchio2025mutants}.
    \item \href{https://github.com/Psivant/femto/tree/main}{Femto} is a Python package for running conventional MD and free energy simulations. Notably it implements tooling for Separated Topologies\cite{baumann2023broadening} and the Alchemical Transfer Method\cite{azimi2022relative,wu2021alchemical,gallicchio2008,gallicchio2025mutants}.

@@

    Konnektor: Tooling to aid in the creation of alchemical perrturbation networks~\cite{ries2024konnektor}.

@@

    \href{https://github.com/cbc-univie/transformato}{transformato} is a package for setting up single topology alchemical transformations to be run using CHARMM or OpenMM \cite{karwounopoulos2022relative, karwounopoulos2023calculations, wieder2022alchemical}.

@@

	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}~\cite{daviddotson2020alchemistry}.
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper      
	\url{https://github.com/choderalab/pymbar} \cite{shirts2008statisticallya}.
	\item \href{https://github.com/OpenFreeEnergy/cinnabar}{Cinnabar}: Standardising alchemical free energy analysis. Includes tools for calculating network-wide analyses and deriving absolute free energies using a maximum-likelihood estimator \cite{richard_gowers_2025_15678719}.
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages 
        \item \href{https://gitlab.com/RutgersLBSR/fe-toolkit}{FE-Toolkit}: collection of packages for free energy analysis, including network-wide MBAR analysis~\cite{giese2025fe}.
    \url{https://github.com/michellab/freenrgworkflows}.
    \item \href{http://www.ties-service.org}{TIES}: Perform statistical analysis of the output of alchemical ensemble simulations and report reliable predictions along with statistical uncertainties for a variety of free energy estimators \url{https://github.com/UCL-CCS/TIES_MD} \cite{bieniek2023}.
	\end{itemize}
\item[] Generally, commercial software will offer more complete pipelines in which standalone analysis applications are not necessarily needed; free and open source packages often require manual analysis but allow more flexibility and modification.
\end{itemize}
